Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-26 overleaf/tech_report/vectorization.tex
Fixed refs and added new related work.

This section describes our vectorization algorithm. While vectorization is a longstanding problem, 
and we build upon existing work on scalar expansion and classical loop vectorization~\cite{Allen:1987}, 
our algorithm is unique as it works on the MPC Source SSA-form representation. We posit that vectorization 
over MPC Source is a new problem that warrants a new look, in part because of MPC's unique linear 
structure and in part because vectorization meshes in with other MPC-specific optimizations in non-trivial ways
(other works have explored manual vectorization and protocol mixing in ad-hoc ways, e.g.,~\cite{Ishaq:2019}).

In this section, we briefly describe the analysis. Details, including edge cases and examples, can be found in the extended version~\cite{Anon_TR}.
%\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
%which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
%\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
%\secref{sec:array_writes} extends vectorization with array writes.

@@

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b) is a same-level edge. 
%A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge. E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
%An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are $\phi$-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\item mixed forward edge $X\rightarrow Y$. $X$ is in some loop $i,j,k$ and $Y$ is in a loop nested into $i,j,k'$. We transform mixed forward edges as follows. Let $\x$ be the variable defined at $X$. We add a variable and assignment $\x' = \x$ immediately after the $i,j,k$ loop. Then we replace the use of $\x$ at $Y$ with $\x'$. This transforms a mixed forward edge into an "inner-to-outer" forward edge followed by an outer-to-inner forward edge. %Thus, Basic Vectorization handles one of "same-level", "inner-to-outer", or "outer-to-inner" def-use edges.
 \end{itemize}

@@

An important component of our algorithm is expansion of scalars and arrays to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
In the Biometric example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to $N*D$ subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scalar {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

Expansion (and also reduction) is done with the $\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ functions, 
which we believe are standard. $\mathit{raise\_dim}$ takes the original array, the access function 
$f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds $((i\!:\!I),(j\!:\!J),(k\!:\!K))$ and produces a new 3-dimensional array 
${\sf A'}$ by iterating over $i,j,k$ and setting each element of  ${\sf A'}$:
\[ \mathit{raise\_dim}(A, f(i,j,k), ((i\!:\!I),(j\!:\!J),(k\!:\!K))): {\sf A'}[i,j,k] =  {\sf A}[f(i,j,k)] \]
Raise dimension applies when reshaping input arrays and also, at outer-to-inner def-use edges.
Analogously $\mathit{drop\_dim}$ applies at inner-to-outer def-use edges.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$. 
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ \mathit{drop\_dim}({\sf A}, (j\!:\!J,k\!:\!K)): {\sf A'}[i] = A[i,J-1,K-1] \] 
%It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
%\[ {\sf A'}[i] = A[i,J-1,K-1] \]

\begin{comment}
\paragraph{Arrays} Conceptually, we treat all variables as arrays. There are three kinds of arrays. 
%\ana{Arrays is not a good name here.}

@@

\paragraph{Drop dimension}  
The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A}, (j:J,k:K)) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'}[i] = A[i,J-1,K-1] \]
\end{comment}

@@

%\subsubsection{Algorithm}

There are two key phases of the algorithm. Phase 1 inserts raise dimension and drop dimension 
operations according to def-uses. E.g., if there is an inner-to-outer dependence, it inserts $\mathit{raise\_dim}$,
and similarly, if there is an outer-to-inner dependence, it inserts $\mathit{drop\_dim}$. 
For example, lines 4,5 in Biometric~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(c) reshape the input arrays. 
After this phase, operations work on arrays of the corresponding dimensionality and we optimistically vectorize all arrays.

Phase 2 proceeds from the inner-most towards the outer-most loop. For each loop it anchors dependence cycles (closures) 
around pseudo PHI nodes then removes vectorization from the dimension of that loop. In the 
Biometric example~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(b), the iteration on the inner loop (7-11)
determines that $j$ is not vectorizable. The next iteration on the outer loop determines that $i$ is vectorizable. It also 
splits the remainder of the loop anchoring each loop around the corresponding PHI node.

There are two important points in this 
phase. First, it may break a loop into smaller loops which would discover opportunities 
for vectorization in intermediate statements in the loop. Second, it handles writes arrays. 
It creates opportunities for vectorization in the presence of write arrays, even though 
Cytron's SSA adds a back-edge to the array PHI-node, thus killing vectorization of statements
that read and write that array.

@@

Phases 3 reverses scalar expansion (this is an optional phase and our current implementation does not include it).
%and Phase 4 explicitly turns operations into MOTION SIMD operations.

\paragraph{Pseudocode:}

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create \python{for} $i_d$ \python{in} ... loop} \COMMENT{ i.e., MOTION loop }
\STATE {add PHI-nodes in $cl$ to header block} 
\STATE {{\color{red}{add target-less PHI-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat \python{for}-loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less PHI-node {\sf A!1 = PHI(A!0,A!k)}}
\STATE {in vectorizable stmts, replace use of {\sf A!1} with {\sf A!0}}
\STATE {discard PHI-node if not used in any $cl$, replacing {\sf A!1} with {\sf A!0} or {\sf A!k} appropriately}
\ENDFOR
}
\ENDFOR

@@

%\COMMENT { Phase 4: }
%\STATE {add SIMD for simdified dimensions}

@@

@@

We build a correctness argument that loosely follows Abstract Interpretation.
First we define the MPC Source syntax. The domain of MPC Source programs
expressible in the syntax (with certain semantic restrictions) is the abstract domain $A$.
We then define the \emph{linearization} of an MPC Source program as an interpretation
over the syntax. The linearization, which is a \emph{schedule} (as in~\secref{sec:overview}), is the concrete domain $C$.
Since we reason over def-use graphs in $A$ we define a partial order relation over elements of $A$
in terms of def-use relations. We define a partial order over elements of $C$ as well,
in terms of def-use relations in the concrete domain $C$. The theorems state 
(informally) that the vectorized program preserves def-uses and thus computes the same 
result as the original program.

\squeeze
\paragraph{MPC Source Syntax}

\figref{fig:mpc_syntax} states the syntax and linearization semantics of MPC Source. 
Although notation is heavy, the linearization simply produces schedules as discussed in~\secref{sec:model}; 
the iterative MPC Source gives rise to what we called sequential schedule where loops are unrolled 
and MPC Source with vectorized dimensions gives rise to what we called parallel schedule. 
For simplicity, we consider only scalars and read-only arrays, however, the treatment extends to write arrays as well.
$\x[i,J,k]$ denotes the value of scalar variable \x{} at loop nest $i,j,k$. 
%(We use 3 dimensions, $i,j,k$, however, the treatment generalizes to arbitrary dimensions.) 
Upper case $J$ denotes a vectorized dimension and lower case $i,k$ denote iterative dimensions. Our compiler imposes 
semantic restrictions over the syntax: (1) $\x$ is treated as a 3-dimensional array and (2) $\x[i,J,k]$ must be enclosed
into for-loops on non-vectorized dimensions $i$ and $k$:
\begin{pythonn}
for i in range(I):
   for k in range(K):
      ... x[i,J,k] ...
\end{pythonn}

@@

\squeeze
\paragraph{Partial Orders}

@@

\squeeze 
\paragraph{Theorems} The two theorems arising from the Basic vectorization optimization are as follows:

@@

These theorems simply state that the transformation preserves def-use relations which is an invariant of the algorithm.

%Theorem 1 states that a def-use edge in MPC Source $a_0$ gives rise to a set of def-use edges in $\gamma(a_0)$
%and a ``same'' edge in $a_1$ gives rise to a set of def-use edges in $\gamma(a_1)$ and the former set is included in the latter one.
%This means that the computation of $\gamma(a_0)$ is carried out by $\gamma(a_1)$ as well, computing the same result. 
%The theorem allows that $a_1$ and $\gamma(a_1)$ have larger state that $a_0$ and $\gamma(a_0)$, i.e., doing additional 
%computation into additional memory locations. Theorem 2 establishes the correctness of vectorization --- 
%clearly, vectorization preserves all statements and def-use edges in the original MPC Source, and therefore the linearization computes the 
%same result as the original MPC Source. (In this case we have an equivalence, $a_0 = a_1$ according to our partial order.)

@@

%\subsubsection{Removal of Infeasible Edges}

Array writes limit vectorization as they sometimes introduce infeasible loop-carried dependencies. For example, consider 
one of our benchmarks, Histogram. The source takes two arrays {\sf A} and {\sf B} of size {\sf N}, where ${\sf A}[j]$ stores the bin number and
${\sf B}[j]$ stores the corresponding value that we need to add. The MPC Source is:

{\small
\begin{pythonn}
for i in range(0, num_bins):
  res1 = PHI(res, res2) # res is the result array
  for j in range(0, N):
    res2 = PHI(res1, res3)
    tmp1 = (A[j] == i)
    tmp2 = (res2[i] + B[j])
    tmp3 = MUX(tmp1, res2[i], tmp2)
    res3 = Update(res2, i, tmp3)
return res1
\end{pythonn}
}

There is a def-use edge to 2 due to the update at 8 of array {\sf res} even though the loop can be vectorized across $i$.

\begin{comment}
Consider the classical example from~\cite{Aiken:1988}:

@@

There is a cycle around {\sf B\_1 = PHI(B\_0,B\_2)} that includes statement
{\sf A\_1 = update(A\_0, i, B\_1[i] + 10)} and that statement won't be vectorized even
though in fact there is no loop-carried dependency from the write of {\sf B\_1[i]} at 7 to the
read of {\sf ... = B\_1[i]} at 6.
\end{comment}

The following algorithm removes certain infeasible loop-carried dependencies that are due to array writes. Consider a loop with index $0 \le j < J$
nested at $i,j,k$. 
%Here $i$ represents the enclosing loops of $j$ and $k$ represents the enclosed loops in $j$.
\begin{comment}
\begin{algorithmic}
\FOR {each array ${\sf A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: ${\sf A}_m[f(i,j,k)] = ... $, and use: $ ... = {\sf A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into ${\sf A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}
\end{comment}
Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) ${\sf A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = {\sf A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for ${\sf A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm, shown in~\cite{Anon_TR}, examines all def-use pairs in loop $j$, including defs and uses in nested loops, 
searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$ (using Z3). If such values exist for some def-use pair, then there is a potential
loop-carried dependence on ${\sf A}$; otherwise there is not and we can remove the spurious backward edge to the array PHI-node 
for loop $j$, thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

In Histogram, this removes the back edge from 4 to the PHI node at 2. Clearly there does not exist $i < i'$ that makes
$i$ at 9 and $i'$ at 6 and 7 equal. The back-edge from 9 to 4 stays because for every $j < j'$ 9 and 6 and 7 access the same location in {\sf res}. %\ana{Can't wrap my head around this...}

\begin{comment}
Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$
due to the def-use pairs of {\sf A} 6-7 and 6-8.
Therefore, we remove the back edge from 6 to the phi-node 2. Analogously, we remove the back edges from 7 to 3 and from 8 to 4. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 9 to 5 remains (def-use pairs for {\sf D}). As a result of removing these spurious edges, Vectorization will find that statement 6 is vectorizable. Statements 7, 8 and 9 will correctly appear in the FOR loop.
\end{comment}

Removing infeasible edges renders some array phi-nodes \emph{target-less}. We handle target-less phi-nodes with a minor extension of Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array {\sf A} back to the closure that updates {\sf A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on {\sf A}, {\sf A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.
Additional detail on the minor changes to $\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ required to do expansion of arrays appear in~\cite{Anon_TR}.

@@

\begin{comment}
\subsubsection{Restricting Array Writes}

@@

\end{comment}

\begin{comment}
\subsubsection{Examples with Array Writes}

@@

\end{comment}

@@

\begin{comment}
\paragraph{Example 2} Now consider the MPC Source of Histogram:

{\small
\begin{pythonn}
for i in range(0, num_bins):
  res1 = PHI(res, res2)
  for j in range(0, N):
    res2 = PHI(res1, res3)
    tmp1 = (A[j] == i)
    tmp2 = (res2[i] + B[j])
    tmp3 = MUX(tmp1, res2[i], tmp2)
    res3 = Update(res2, i, tmp3)
return res1
\end{pythonn}
}
\end{comment}

Returning to the histogram example, the canonical (i.e., declared) dimensionality of {\sf res} is 1. Also, the phi-node {\sf res1 = PHI(res, res2)} is a target-less phi-node (the implication being that the inner for loop can be vectorized across $i$). After Phase 1, Vectorization produces the following code (statements are implicitly vectorized along $i$ and $j$). In a vectorized update statement, we can ignore the incoming array, {\sf res2} in this case. The update writes (in parallel) all locations of the 2-d array, in this case it sets up each {\sf res3[i,j] = tmp3[i,j]}.

@@

Processing the inner loop in Phase 2 vectorizes {\sf tmp1 = (A1 == I)} along the $j$ dimension but leaves the rest of the statements in a MOTION loop. Processing the outer loop is interesting. This is because the PHI node is a target-less node, and therefore, there are no closures. (1) Everything can be vectorized along the $i$ dimension. (2) We remove the target-less PHI node, however, we must update uses of {\sf res1} appropriately: the use at $\mathit{raise\_dim}$ goes to the first argument of the PHI function and the use at $\mathit{drop\_dim}$ goes to the second argument.

@@

res1^ = raise_dim(res, (j:N)) # replacing res1 with res, 1st arg
for j in range(0, N):
   res2 = PHI(res1^, res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
   equiv. to res3 = res2; res3[I,j] = tmp3[I,j]
res2^ = drop_dim(res2)
res1 = drop_dim(res2^) # replacing with res2^, 2nd arg. NOOP
return res1
\end{pythonn}
}

>>>>

This section describes our vectorization algorithm. While vectorization is a longstanding problem, 
and we build upon existing work on scalar expansion and classical loop vectorization~\cite{Allen:1987}, 
our algorithm is unique as it works on the MPC Source SSA-form representation. We posit that vectorization 
over MPC Source is a new problem that warrants a new look, in part because of MPC's unique linear 
structure and in part because vectorization meshes in with other MPC-specific optimizations in non-trivial ways
(other works have explored manual vectorization and protocol mixing in an ad-hoc way, e.g.,~\cite{Demmler:2015,Buscher:2018,Ishaq:2019}).

\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
\secref{sec:array_writes} extends vectorization with array writes.

@@

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source} is a same-level edge. A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge.
E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are $\phi$-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\item mixed forward edge $X\rightarrow Y$. $X$ is in some loop $i,j,k$ and $Y$ is in a loop nested into $i,j,k'$. We transform mixed forward edges as follows. Let $\x$ be the variable defined at $X$. We add a variable and assignment $\x' = \x$ immediately after the $i,j,k$ loop. Then we replace the use of $\x$ at $Y$ with $\x'$. This transforms a mixed forward edge into an "inner-to-outer" forward edge followed by an outer-to-inner forward edge. Thus, Basic Vectorization handles one of "same-level", "inner-to-outer", or "outer-to-inner" def-use edges.
 \end{itemize}

@@

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
In the Biometric example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to $N*D$ subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scalar {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

\paragraph{Arrays} Conceptually, we treat all variables as arrays. There are three kinds of arrays. 
%\ana{Arrays is not a good name here.}

@@

\paragraph{Drop dimension}  
The corresponding $\mathit{drop\_dim}$ is carried out when an array written in an inner loop is used in an enclosing loop.
It takes a higher dimensional array, say $i,j,k$ and removes trailing dimensions, say $j,k$:
\[ \mathit{drop\_dim}({\sf A}, (j:J,k:K)) \]
It iterates over $i$ and takes the result at the maximal index of $j$ and $k$, i.e., the result at the last iterations of $j$ and $k$:
\[ {\sf A'}[i] = A[i,J-1,K-1] \]

@@

We present our algorithm followed by an example on Biometric.

\subsubsection{Algorithm}

There are two key phases of the algorithm. Phase 1 inserts raise dimension and drop dimension 
operations according to def-uses. E.g., if there is an inner-to-outer dependence, it inserts $\mathit{raise\_dim}$,
and similarly, if there is an outer-to-inner dependence, it inserts $\mathit{drop\_dim}$. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays.

Phase 2 proceeds from the inner-most towards the outer-most loop. For each loop it anchors dependence cycles (closures) 
around pseudo PHI nodes then removes vectorization from the dimension of that loop. There are two important points in this 
phase. First, it may break a loop into smaller loops which would discover opportunities 
for vectorization in intermediate statements in the loop. Second, it handles writes arrays. 
It creates opportunities for vectorization in the presence of write arrays, even though 
Cytron's SSA adds a back-edge to the array PHI-node, thus killing vectorization of statements
that read and write that array.

@@

Phases 3 cleans up local arrays of references (this is an optional phase and our current implementation does not include it)
and Phase 4 explicitly turns operations into MOTION SIMD operations.

@@

\WHILE {there are closure $cl_1$ and $cl_2$ that intersect}
\STATE {merge $cl_1$ and $cl_2$}
\ENDWHILE
\FOR {each closure $cl$ (after merge)}
\STATE {create \python{for} $i_d$ \python{in} ... loop}
\STATE {add PHI-nodes in $cl$ to header block}
\STATE {{\color{red}{add target-less PHI-node for A if $cl$ updates array A}}}
\STATE {add statements in $cl$ to loop in some order of dependences}
\STATE \COMMENT { Dimension is not vectorizable: }
\STATE {change ${I_d}$ to $i_d$ in all statements in loop}
\STATE {treat \python{for}-loop as monolith node for def-uses: e.g., some def-use edges become same-level.}
\ENDFOR
{
\color{red}
\FOR {each target-less PHI-node {\sf A!1 = PHI(A!0,A!k)}}
\STATE {in vectorizable stmts, replace use of {\sf A!1} with {\sf A!0}}
\STATE {discard PHI-node if not used in any $cl$, replacing {\sf A!1} with {\sf A!0} or {\sf A!k} appropriately}
\ENDFOR
}
\ENDFOR

@@

\COMMENT { Phase 4: }
\STATE {add SIMD for simdified dimensions}

@@

We build a correctness argument that loosely follows Abstract Interpretation.
First we define the MPC Source syntax. The domain of MPC Source programs
expressible in the syntax (with certain semantic restrictions) is the abstract domain $A$.
We then define the \emph{linearization} of an MPC Source program as an interpretation
over the syntax. The linearization, which is a \emph{schedule} (as in~\secref{sec:overview}), is the concrete domain $C$.
Since we reason over def-use graphs in $A$ we define a partial order relation over elements of $A$
in terms of def-use relations. We define a partial order over elements of $C$ as well,
in terms of def-use relations in the concrete domain $C$. We prove two theorems
that state (informally) that the schedule corresponding to the original program computes the same
result as the schedule corresponding to the vectorized program.

@@

\paragraph{MPC Source Syntax}

\figref{fig:mpc_syntax} states the syntax and linearization semantics of MPC Source. 
Although notation is heavy, the linearization simply produces schedules as discussed in~\secref{sec:model}; 
the iterative MPC Source gives rise to what we called sequential schedule where loops are unrolled 
and MPC Source with vectorized dimensions gives rise to what we called parallel schedule. 
For simplicity, we consider only scalars and read-only arrays, however, the treatment extends to write arrays as well.
$\x[i,J,k]$ denotes the value of scalar variable \x{} at loop nest $i,j,k$. 
%(We use 3 dimensions, $i,j,k$, however, the treatment generalizes to arbitrary dimensions.) 
Upper case $J$ denotes a vectorized dimension and lower case $i,k$ denote iterative dimensions. Our compiler imposes 
semantic restrictions over the syntax: (1) $\x$ is treated as a 3-dimensional array and (2) $\x[i,J,k]$ must be enclosed
into for-loops on non-vectorized dimensions $i$ and $k$:
\begin{pythonn}
for i in range(I):
   ...
   for k in range(K):
      ... x[i,J,k] ...
\end{pythonn}

@@

\paragraph{Partial Orders}

@@

\paragraph{Theorems} The two theorems arising from the Basic vectorization optimization are as follows:

@@

Theorem 1 states that a def-use edge in MPC Source $a_0$ gives rise to a set of def-use edges in $\gamma(a_0)$
and a ``same'' edge in $a_1$ gives rise to a set of def-use edges in $\gamma(a_1)$ and the former set is included in the latter one.
This means that the computation of $\gamma(a_0)$ is carried out by $\gamma(a_1)$ as well, computing the same result. 
The theorem allows that $a_1$ and $\gamma(a_1)$ have larger state that $a_0$ and $\gamma(a_0)$, i.e., doing additional 
computation into additional memory locations. Theorem 2 establishes the correctness of vectorization --- 
clearly, vectorization preserves all statements and def-use edges in the original MPC Source, and therefore the linearization computes the 
same result as the original MPC Source. (In this case we have an equivalence, $a_0 = a_1$ according to our partial order.)

@@

\subsubsection{Removal of Infeasible Edges}

Array writes limit vectorization as they sometimes introduce infeasible loop-carried dependencies. Consider the classical example from~\cite{Aiken:1988}:

@@

There is a cycle around {\sf B\_1 = PHI(B\_0,B\_2)} that includes statement
{\sf A\_1 = update(A\_0, i, B\_1[i] + 10)} and that statement won't be vectorized even
though in fact there is no loop-carried dependency from the write of {\sf B\_1[i]} at 7 to the
read of {\sf ... = B\_1[i]} at 6.

The following algorithm removes certain infeasible loop-carried dependencies that are due to array writes. Consider a loop with index $0 \le j < J$
nested at $i,j,k$. Here $i$ represents the enclosing loops of $j$ and $k$ represents the enclosed loops in $j$.

\begin{algorithmic}
\FOR {each array ${\sf A}$ written in loop $j$}
\STATE \COMMENT { including enclosed loops in $j$ }
\STATE dep = False
\FOR {each pair def: ${\sf A}_m[f(i,j,k)] = ... $, and use: $ ... = {\sf A}_n[f'(i,j,k)]$ in loop $j$}
\IF {$\exists \underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$, s.t. $0\! \le\! \underline{i}\! <\! I,\: 0\! \le\! \underline{j}, \underline{j}' \! <\! J,\: 0\! \le\! \underline{k}, \underline{k}' \! <\! K,\: \underline{j}<\underline{j}'$, and $f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$}
\STATE dep = True
\ENDIF
\ENDFOR
\IF {dep == False}
\STATE remove back edge into ${\sf A}$'s $\phi$-node in loop $j$.
\ENDIF
\ENDFOR
\end{algorithmic}

Consider a loop $j$ enclosed in some fixed $\underline{i}$. Only if an update (definition) ${\sf A}_m[f(i,j,k)] = ... $ at some iteration $\underline{j}$
references the \emph{same} array element as a use $ ... = {\sf A}_n[f'(i,j,k)]$ at some later iteration $\underline{j}'$,
we may have a loop-carried dependence for ${\sf A}$ due to this def-use pair. (In contrast, Cytron's algorithm inserts a loop-carried dependency every time there is an array update.)
The algorithm above examines all def-use pairs in loop $j$, including defs and uses in nested loops, searching for values $\underline{i}, \underline{j}, \underline{j}', \underline{k}, \underline{k}'$ that satisfy
$f(\underline{i},\underline{j},\underline{k}) = f'(\underline{i},\underline{j}',\underline{k}')$. If such values exist for some def-use pair, then there is a potential
loop-carried dependence on ${\sf A}$; otherwise there is not and we can remove the spurious backward edge thus ``freeing up'' statements for vectorization. %\ana{Double check this!!!}

Consider the earlier example. There is a single loop, $i$. Clearly, there is no pair $\underline{i}$ and $\underline{i}'$, where $\underline{i} < \underline{i}'$ that make $\underline{i} = \underline{i}'$
due to the def-use pairs of {\sf A} 6-7 and 6-8.
Therefore, we remove the back edge from 6 to the phi-node 2. Analogously, we remove the back edges from 7 to 3 and from 8 to 4. However, there are many values $\underline{i} < \underline{i'}$ that make $\underline{i} = \underline{i'}-1$ and the back edge from 9 to 5 remains (def-use pairs for {\sf D}). As a result of removing these spurious edges, Vectorization will find that statement 6 is vectorizable. Statements 7, 8 and 9 will correctly appear in the FOR loop.

Note however, that this step renders some array phi-nodes target-less. We handle target-less phi-nodes with a minor extension of Vectorization (Phase 2).
First, we merge closures that update the same array. This simplifies handling of array $\phi$-nodes: if each closure is turned into a separate loop
each loop will need to have its own array phi-node to account for the update and this would complicate the analysis.
Second, we add the target-less node of array {\sf A} back to the closure that updates {\sf A} ---
the intuition is, even if there is no loop-carried dependence from writes to reads on {\sf A}, {\sf A} is written and the write (i.e., update) cannot be vectorized;
therefore, the updated array has to carry to the next iteration of the loop. Third, in cases when the phi-node remains target-less, i.e., cases when the
array write can be vectorized, we have to properly remove the phi-node replacing uses of the left-hand side of the phi-node with its arguments.

@@

\subsubsection{Restricting Array Writes}

@@

\subsubsection{Examples with Array Writes}

@@

@@

\paragraph{Example 2} Now consider the MPC Source of Histogram:

{\small
\begin{pythonn}
for i in range(0, num_bins):
  res1 = PHI(res, res2)
  for j in range(0, N):
    res2 = PHI(res1, res3)
    tmp1 = (A[j] == i)
    tmp2 = (res2[i] + B[j])
    tmp3 = MUX(tmp1, res2[i], tmp2)
    res3 = Update(res2, i, tmp3)
return res1
\end{pythonn}
}

The canonical dimensionality of {\sf res} is 1. Also, the phi-node {\sf res1 = PHI(res, res2)} is a target-less phi-node (the implication being that the inner for loop can be vectorized across $i$). After Phase 1, Vectorization produces the following code (statements are implicitly vectorized along $i$ and $j$). In a vectorized update statement, we can ignore the incoming array, {\sf res2} in this case. The update writes (in parallel) all locations of the 2-dimensional array, in this case it sets up each {\sf res3[i,j] = tmp3[i,j]}.

@@

Processing the inner loop in Phase 2 vectorizes {\sf tmp1 = (A1 == I)} along the $j$ dimension but leaves the rest of the statements in a MOTION loop. Processing the outer loop is interesting. This is because the PHI node is a target-less node, and therefore, there are no closures! Several things happen. (1) Everything can be vectorized along the $i$ dimension. (2) We remove the target-less PHI node, however, we must update uses of {\sf res1} appropriately: the use at $\mathit{raise\_dim}$ goes to the first argument of the PHI function and the use at $\mathit{drop\_dim}$ goes to the second argument.

@@

res1^ = raise_dim(res, (j:N)) // replacing res1 with res, 1st arg
for j in range(0, N):
   res2 = PHI(res1^, res3)
   tmp2[I,j] = (res2[I,j] + B1[I,j])
   tmp3[I,j] = MUX(tmp1[I,j], res2[I,j], tmp2[I,j])
   res3 = Update(res2, (I,j), tmp3)
   equiv. to res3 = res2; res3[I,j] = tmp3[I,j]
res2^ = drop_dim(res2)
res1 = drop_dim(res2^) // replacing with res2^, 2nd arg. NOOP
return res1
\end{pythonn}
}
